A Generic and Efficient E-field Parallel Imaging Correlator for Next-Generation Radio Telescopes

### 2016-08-24 imaging_demo/ms.tex
Shorten section 5, add appendix for detailed math
From 601fdcb (Judd-v1):
======================

In ms.tex,

(*) Shorten section 5 and add appendix for detailed math

A high level software architecture of the EPIC package is described in the 
appendix \S\ref{sec:software-modules} for the interested reader.

@@

{\bf One of the advanced features of the MOFF algorithm and its implementation via EPIC is the ability to naturally account for heterogeneity of antennas while still producing images with maximal information \citep{mor11}. Unlike the assumption frequently made in standard radio interferometry that all antennas are identical, EPIC can implement the most generic case where no such assumption can be made. In other words, EPIC does not assume $W_a(\mathbf{r}-\mathbf{r}_a)$ or $B_{ab}(\mathbf{r}-\mathbf{r}_{ab})$ for different antennas are identical. However, any {\it a priori} knowledge available that certain sets of antennas have identical illumination patterns can be easily passed on to EPIC to avoid redundant computations and thus increase efficiency.

We present a methodology to understand the effective angular weighting in the image obtained with EPIC from a heterogeneous array and compare it with that from standard interferometry which assumes a homogeneous array (all antennas are identical). We also demonstrate how the assumption in the latter results in a mis-estimation of the recovered sky model.

Dropping the noise term and time dependence, the visibility measured by an antenna pair can be written from equation~\ref{eqn:measurement-eqn-1} as:
\begin{align}\label{eqn:measurement-eqn-2}
  V_{ab}(f) &= \int \mathcal{W}^\textrm{I}_a(\hat{\mathbf{s}},f)\mathcal{W}^{\textrm{I}\star}_b(\hat{\mathbf{s}},f)\,\mathcal{I}(\hat{\mathbf{s}},f)\,e^{-2\pi i f\mathbf{r}_{ab}\!\cdot\,\hat{\mathbf{s}}/c}\,\dif\Omega,
\end{align}
where, superscript `$\textrm{I}$' in the $\mathcal{W}$ term indicates the weighting is inherently introduced by the instrument during the measurement process. The beam-weighted dirty image is:
\begin{align}
  \mathcal{I}_\textrm{D}(\hat{\mathbf{s}},f) &= \frac{1}{\Nant(\Nant-1)/2}\,\sum_{ab}\, V_{ab}(f)\,e^{2\pi i f\mathbf{r}_{ab}\!\cdot\,\hat{\mathbf{s}}/c},
\end{align}
{\bf where, we have assumed equal weighting for each antenna pair}.

When the array is homogeneous, $\mathcal{W}^\textrm{I} \equiv \mathcal{W}^\textrm{I}_a \equiv \mathcal{W}^\textrm{I}_b$ and the above equation reduces to:
\begin{align}\label{eqn:dirty-image-homogeneous}
  \mathcal{I}_\textrm{D}(\hat{\mathbf{s}},f) &= |\mathcal{W}^\textrm{I}(\hat{\mathbf{s}},f)|^2\,\mathcal{I}_\textrm{D}^\textrm{iso}(\hat{\mathbf{s}},f),
\end{align}
where, $\mathcal{I}_\textrm{D}^\textrm{iso}(\hat{\mathbf{s}},f)$ is the dirty image with no beam weighting (isotropic, uniform weighting {\bf of the sky}) determined only by the array layout and $|\mathcal{W}^\textrm{I}|^2$ is the directional power pattern of the antenna pair familiar in standard interferometry.

{\bf In order to understand an image from a heterogeneous array, we start by looking at the contribution to the image from the visibility, $V_{ab}$, from each antenna pair.} While imaging, EPIC introduces a weighting to each antenna during gridding as given by equations~\ref{eqn:dirty-image-FX} and \ref{eqn:e-field-conv}, and equivalently, the weighted visibility from {\bf an} antenna pair $ab$ projected on the grid is:
\begin{align}
  V^\prime_{ab}(\mathbf{r},f) &= B^{\textrm{G}\star}_{ab}(\mathbf{r}-\mathbf{r}_{ab},f)\,V_{ab},
\end{align}
where, $B^\textrm{G}_{ab}(\mathbf{r}-\mathbf{r}_{ab},f)$ is obtained by the spatial cross-correlation of weighting kernels associated with the individual antennas using the definition:
\begin{align}
  B^\textrm{G}_{ab}(\mathbf{r}-\mathbf{r}_{ab},f) &= \int W^\textrm{G}_a(\mathbf{r}^\prime-\mathbf{r}_a)\,W^{\textrm{G}\star}_b(\mathbf{r}^\prime+\mathbf{r}-\mathbf{r}_b)\,\dif^2\mathbf{r}^\prime.
\end{align}
The superscript `$\textrm{G}$' denotes the weighting introduced in analysis during the gridding process. Optimal imaging requires $W^\textrm{G}_a = W^\textrm{I}_a$ \citep{mor09,mor11}. However, we keep the two superscripts separately to describe output images under the wrong assumption that the array is homogeneous when instead it is actually heterogeneous.

The sky response of the weighted visibility is:
\begin{align}
  \mathcal{I}^\prime_{ab}(\hat{\mathbf{s}},f) &= \int V^\prime_{ab}(\mathbf{r},f)\,e^{2\pi i f\mathbf{r}\cdot\,\hat{\mathbf{s}}/c}\,\dif^2\mathbf{r} \nonumber\\
                                              &= \int B^{\textrm{G}\star}_{ab}(\mathbf{r}-\mathbf{r}_{ab},f)\,V_{ab}\,e^{2\pi i f\mathbf{r}\cdot\,\hat{\mathbf{s}}/c}\,\dif^2\mathbf{r} \nonumber\\
                                              &= V_{ab}\,\int \left[\int W^{\textrm{G}\star}_a(\mathbf{r}^\prime-\mathbf{r}_a)\,W^\textrm{G}_b(\mathbf{r}^\prime+\mathbf{r}-\mathbf{r}_a)\,\dif^2\mathbf{r}^\prime\right] \nonumber\\
                                              &\qquad\qquad \times e^{2\pi i f\mathbf{r}\cdot\,\hat{\mathbf{s}}/c}\,\dif^2\mathbf{r} \nonumber\\
                                              &= \mathcal{W}^{\textrm{G}\star}_a(\hat{\mathbf{s}},f)\mathcal{W}^\textrm{G}_b(\hat{\mathbf{s}},f)\,V_{ab}\,e^{2\pi i f\mathbf{r}_{ab}\!\cdot\,\hat{\mathbf{s}}/c},
\end{align}
which denotes the fringe from the weighted visibility. The image output of EPIC is equivalent to averaging these weighted fringes from all antenna pairs:
\begin{align}\label{eqn:wt-dirty-image-EPIC}
  \mathcal{I}^\prime(\hat{\mathbf{s}},f) &= \frac{1}{\Nant(\Nant-1)/2}\,\sum_{\substack{ab\\a\ne b}}\,\mathcal{I}^\prime_{ab}(\hat{\mathbf{s}},f) \nonumber\\
                                         &= \frac{1}{\Nant(\Nant-1)/2}\,\sum_{\substack{ab\\a\ne b}}\,\left[\mathcal{W}^{\textrm{G}\star}_a(\hat{\mathbf{s}},f)\mathcal{W}^\textrm{G}_b(\hat{\mathbf{s}},f)\,V_{ab}\right. \nonumber\\
  &\qquad\qquad\qquad\qquad\qquad\qquad \times \left. e^{2\pi i f\mathbf{r}_{ab}\!\cdot\,\hat{\mathbf{s}}/c}\right].
\end{align}

If all antennas are identically weighted ($W^\textrm{G} \equiv W^\textrm{G}_a \equiv W^\textrm{G}_b$), using equation~\ref{eqn:dirty-image-homogeneous}, equation~\ref{eqn:wt-dirty-image-EPIC} reduces to: 
\begin{align}\label{eqn:wt-dirty-image-homogeneous}
  \mathcal{I}^\prime(\hat{\mathbf{s}},f) &= \frac{\left|\mathcal{W}^\textrm{G}(\hat{\mathbf{s}},f)\right|^2}{\Nant(\Nant-1)/2}\,\sum_{\substack{ab\\a\ne b}}\,V_{ab}\,e^{2\pi i f\mathbf{r}_{ab}\!\cdot\,\hat{\mathbf{s}}/c} \nonumber\\
  &= \left|\mathcal{W}^\textrm{G}(\hat{\mathbf{s}},f)\right|^2\,\mathcal{I}_\textrm{D}(\hat{\mathbf{s}},f) \nonumber\\
  &= \left|\mathcal{W}^\textrm{G}(\hat{\mathbf{s}},f)\right|^2\,\left|\mathcal{W}^\textrm{I}(\hat{\mathbf{s}},f)\right|^2\,\mathcal{I}_\textrm{D}^\textrm{iso}(\hat{\mathbf{s}},f).
\end{align}
Thus, the dirty image, which is already attenuated by the instrumental power pattern, gets further attenuated by the power pattern introduced in the gridding step in EPIC imager. This is consistent with \citet{mor09}.

{\bf We note that all quantities in the celestial plane (calligraphic fonts) are functions of position, $\hat{\mathbf{s}}$, and frequency, $f$. For convenience, we drop writing this dependence explicitly hereafter.}

{\bf For a heterogeneous array, it is necessary to keep track of antennas and antenna pairs of different types which will determine the final weighting in the synthesized image.} We consider a heterogeneous array consisting of a total of $\Nant$ antennas under $N_\textrm{T}$ different types. And under each antenna type $i$, there are $n_i$ antennas such that $\sum_i n_i = \Nant$. The total number of unique non-zero spacing antenna pairs is $\Nant(\Nant-1)/2$ and there are potentially up to $N_\textrm{T}(N_\textrm{T}+1)/2$ unique antenna pair types obtained by pairwise combination of the $N_\textrm{T}$ antenna types. Thus, the total number of antenna pairs, 
\begin{align}\label{eqn:total-baselines}
  \frac{1}{2}\Nant(\Nant-1) &= \frac{1}{2}\,\sum_{ij} n_{ij}, \\
  \textrm{where}, \quad   n_{ij} &= \begin{cases}
    n_i\,n_j, & i\ne i \\
    n_i(n_i-1), & i=j, 
  \end{cases}
\end{align}
is obtained by counting antenna pairs, $n_{ij}$, in different antenna pair types, $ij$.

Since equation~\ref{eqn:wt-dirty-image-EPIC} is obtained by averaging weighted fringes over all unique antenna pairs, it can be expressed as: 
\begin{align}\label{eqn:wt-dirty-image-EPIC-decomp}
  \mathcal{I}^\prime &= \frac{1/2}{\Nant(\Nant-1)/2}\,\sum_{ij}\sum_{\substack{ab\\a\ne b\\a\in i, b\in j}}\,\mathcal{I}^\prime_{ab}
\end{align}
Here, equation~\ref{eqn:wt-dirty-image-EPIC} has been decomposed as the sum of fringes over all unique antenna pairs in an antenna pair type (inner sum) and subsequently summed over all antenna pair types (outer sum).

For purposes of demonstration in this study, we make the following simplification. We assume that the spatial distribution of antenna pairs in each antenna pair type are similar and thus result in a similar dirty image, $\mathcal{I}_\textrm{D}^\textrm{iso}$. This usually holds when $\Nant$ is large and antennas under each antenna type are chosen randomly such that differences in the point spread functions are insignificant. Under this assumption, equation~\ref{eqn:wt-dirty-image-EPIC-decomp} can be written as:
\begin{align}\label{eqn:wt-dirty-image-EPIC-decomp-approx1}
  \mathcal{I}^\prime &\approx \frac{\mathcal{I}_\textrm{D}^\textrm{iso}}{\Nant(\Nant-1)}\,\sum_{ij}\sum_{\substack{ab\\a\ne b\\a\in i, b\in j}} \mathcal{W}^{\textrm{G}\star}_a\,\mathcal{W}^\textrm{G}_b\,\mathcal{W}^\textrm{I}_a\,\mathcal{W}^{\textrm{I}\star}_b,
\end{align}
where, $\mathcal{I}_\textrm{D}^\textrm{iso}$ is approximately the same for all terms and has been pulled outside the summations. For all antennas indexed by $a$ that belong to a particular type $i$, we replace $\mathcal{W}_a$ with $\mathcal{W}_{(i)}$ {\bf and is applicable to those arising} from both instrumental (superscript `$I$') and gridding (superscript `$G$') origins. Using arguments similar to those in equation~\ref{eqn:total-baselines}, equation~\ref{eqn:wt-dirty-image-EPIC-decomp-approx1} can be further simplified to:
\begin{align}\label{eqn:wt-dirty-image-EPIC-decomp-approx2}
  \mathcal{I}^\prime &\approx \frac{\mathcal{I}_\textrm{D}^\textrm{iso}}{\Nant(\Nant-1)}\,\sum_{ij}n_{ij}\,\mathcal{W}^{\textrm{G}\star}_{(i)}\,\mathcal{W}^\textrm{G}_{(j)}\,\mathcal{W}^\textrm{I}_{(i)}\,\mathcal{W}^{\textrm{I}\star}_{(j)}.
\end{align}
Then the effective attenuation of the dirty image due to instrumental and gridding weights is:
\begin{align}\label{eqn:effective-weighting}
  \mathcal{W}_\textrm{eff} &= \mathcal{I}^\prime\, / \,\mathcal{I}_\textrm{D}^\textrm{iso}.
\end{align}
This means that the effective attenuation of the dirty image is obtained by combinations of the fourth power of antenna voltage patterns each weighted by the number of antenna pairs in those antenna pair types. In an optimally weighted image, $\mathcal{W}^\textrm{G}_{(i)} \equiv \mathcal{W}^\textrm{I}_{(i)}$ and thus, equation~\ref{eqn:effective-weighting} reduces to:
\begin{align}\label{eqn:effective-weighting-optimal}
  \mathcal{W}_\textrm{eff}^\textrm{opt} &\approx \frac{1}{\Nant(\Nant-1)}\,\sum_{ij} n_{ij}\,\left|\mathcal{W}^\textrm{I}_{(i)}\right|^2\left|\mathcal{W}^\textrm{I}_{(j)}\right|^2,
\end{align}
where the superscript `$\textrm{opt}$' denotes `optimal'.

@@

>>>>

A high level software architecture of the EPIC package is described in the 
appendix~\ref{sec:software-modules} for the interested reader.

@@

{\bf One of the advanced features of the MOFF algorithm and its implementation via EPIC is the ability to naturally account for heterogeneity of antennas while still producing images with maximal information \citep{mor11}. Unlike the assumption frequently made in existing direct imaging techniques that all antennas are identical, EPIC can implement the most generic case where no such assumption can be made. In other words, EPIC does not assume $W_a(\mathbf{r}-\mathbf{r}_a)$ or $B_{ab}(\mathbf{r}-\mathbf{r}_{ab})$ for different antennas are identical. However, any {\it a priori} knowledge available that certain sets of antennas have identical illumination patterns can be easily passed on to EPIC to avoid redundant computations and thus increase efficiency. Here we present key highlights from the generic methodology detailed in appendix~\ref{sec:math-versatility} to understand the effective angular weighting in the image obtained with EPIC from a heterogeneous array.

We look at the contribution to the image from the visibility, $V_{ab}$, from each antenna pair. While imaging, EPIC introduces a weighting to each antenna during gridding as given by equations~\ref{eqn:dirty-image-FX} and \ref{eqn:e-field-conv}, and equivalently, the weighted visibility from {\bf an} antenna pair $ab$ projected on the grid is:
\begin{align}
  V^\prime_{ab}(\mathbf{r},f) &= B^{\textrm{G}\star}_{ab}(\mathbf{r}-\mathbf{r}_{ab},f)\,V_{ab},
\end{align}
where, $B^\textrm{G}_{ab}(\mathbf{r}-\mathbf{r}_{ab},f)$ is obtained by the spatial cross-correlation of weighting kernels associated with the individual antennas. The superscript `$\textrm{G}$' denotes the weighting introduced in analysis during the gridding process. We will use superscript `I' to denote the inherent weighting applied by the instrument during the measurement process in obtaining $V_{ab}$ from the true sky model. Optimal imaging requires $W^\textrm{G}_a = W^\textrm{I}_a$ \citep{mor09,mor11}. However, we keep the two superscripts separately to describe output images under the wrong assumption that the array is homogeneous when instead it is actually not.

The sky response of the weighted visibility, referred to as {\it fringes}, is:
\begin{align}
  \mathcal{I}^\prime_{ab}(\hat{\mathbf{s}},f) &= \mathcal{W}^{\textrm{G}\star}_a(\hat{\mathbf{s}},f)\mathcal{W}^\textrm{G}_b(\hat{\mathbf{s}},f)\,V_{ab}\,e^{2\pi i f\mathbf{r}_{ab}\!\cdot\,\hat{\mathbf{s}}/c}.
\end{align}
Since all quantities in the celestial plane (calligraphic fonts) are functions of position, $\hat{\mathbf{s}}$, and frequency, $f$, we drop writing this dependence explicitly hereafter.

The image output of EPIC is the average of these weighted fringes from all antenna pairs:
\begin{align}\label{eqn:wt-dirty-image-EPIC}
  \mathcal{I}^\prime &= \frac{1}{\Nant(\Nant-1)}\,\sum_{\substack{ab\\a\ne b}}\mathcal{I}^\prime_{ab}
\end{align}

We define the effective attenuation of the dirty image due to instrumental and gridding weights as:
\begin{align}\label{eqn:effective-weighting}
  \mathcal{W}_\textrm{eff} &= \mathcal{I}^\prime\, / \,\mathcal{I}_\textrm{D}^\textrm{iso},
\end{align}
where, $\mathcal{I}_\textrm{D}^\textrm{iso}$ is the dirty image with no beam weighting (isotropic, uniform weighting of the sky) determined only by the array layout.

For a homogeneous array, $W^\textrm{G} \equiv W^\textrm{G}_a \equiv W^\textrm{G}_b$. Thus,  
\begin{align}\label{eqn:wt-dirty-image-homogeneous}
  \mathcal{I}^\prime &= \left|\mathcal{W}^\textrm{G}\right|^2\,\left|\mathcal{W}^\textrm{I}\right|^2\,\mathcal{I}_\textrm{D}^\textrm{iso}.
\end{align}
Consistent with \citet{mor09}, the dirty image, which is already attenuated by the instrumental power pattern inherent in the measurement process, gets further attenuated by the power pattern introduced in the gridding step in EPIC imager.

For a heterogeneous array, 
\begin{align}\label{eqn:wt-dirty-image-EPIC-decomp}
  \mathcal{I}^\prime &= \frac{1}{\Nant(\Nant-1)}\,\sum_{ij}\sum_{\substack{ab\\a\ne b\\a\in i, b\in j}}\,\mathcal{I}^\prime_{ab},
\end{align}
where, $ab$ indexes pairs of antennas $a$ and $b$, $ij$ indexes types of pairs of antennas, with $i$ and $j$ indexing antenna types. Thus, $\mathcal{I}^\prime$ is written as the sum of fringes over all unique antenna pairs in an antenna pair type (inner sum) and subsequently summed over all antenna pair types (outer sum).

For purposes of demonstration in this study, we assume that the spatial distribution of antenna pairs in each antenna pair type is similar and thus results in a similar dirty image, $\mathcal{I}_\textrm{D}^\textrm{iso}$. This usually holds when antennas under each antenna type are chosen randomly or identically such that differences in the synthesized beams are insignificant. Then, equation~\ref{eqn:wt-dirty-image-EPIC-decomp} can be approximated as:
\begin{align}\label{eqn:wt-dirty-image-EPIC-decomp-approx1}
  \mathcal{I}^\prime &\approx \frac{\mathcal{I}_\textrm{D}^\textrm{iso}}{\Nant(\Nant-1)}\,\sum_{ij}\sum_{\substack{ab\\a\ne b\\a\in i, b\in j}} \mathcal{W}^{\textrm{G}\star}_a\,\mathcal{W}^\textrm{G}_b\,\mathcal{W}^\textrm{I}_a\,\mathcal{W}^{\textrm{I}\star}_b,
\end{align}
where, $\mathcal{I}_\textrm{D}^\textrm{iso}$ is the same for all terms and has been pulled out of the summations. For all antennas indexed by $a$ that belong to a particular type $i$, we replace $\mathcal{W}_a$ with $\mathcal{W}_{(i)}$ {\bf and is applicable to those arising} from both instrumental (superscript `$I$') and gridding (superscript `$G$') origins. If $n_{ij}$ is the number of antenna pairs in the antenna pair type $ij$, $\mathcal{W}_\textrm{eff}$ can be simplified as:
\begin{align}\label{eqn:wt-dirty-image-EPIC-decomp-approx2}
  \mathcal{W}_\textrm{eff} &\approx \frac{1}{\Nant(\Nant-1)}\,\sum_{ij}n_{ij}\,\mathcal{W}^{\textrm{G}\star}_{(i)}\,\mathcal{W}^\textrm{G}_{(j)}\,\mathcal{W}^\textrm{I}_{(i)}\,\mathcal{W}^{\textrm{I}\star}_{(j)}.
\end{align}
In an optimally weighted image, $\mathcal{W}^\textrm{G}_{(i)} \equiv \mathcal{W}^\textrm{I}_{(i)}$. Thus,
\begin{align}\label{eqn:effective-weighting-optimal}
  \mathcal{W}_\textrm{eff}^\textrm{opt} &\approx \frac{1}{\Nant(\Nant-1)}\,\sum_{ij} n_{ij}\,\left|\mathcal{W}^\textrm{I}_{(i)}\right|^2\left|\mathcal{W}^\textrm{I}_{(j)}\right|^2,
\end{align}
where, superscript `$\textrm{opt}$' denotes `optimal'.

@@

\section{Mathematical Framework for Imaging with Heterogeneous Arrays}\label{sec:math-versatility}

{\bf We present a methodology to understand the effective angular weighting in the image obtained with EPIC from a heterogeneous array and compare it with that from existing imaging applications which assume a homogeneous array (all antennas are identical). % We also demonstrate how the assumption in the latter results in a mis-estimation of the recovered sky model.

Dropping the noise term and time dependence, the visibility measured by an antenna pair can be written from equation~\ref{eqn:measurement-eqn-1} as:
\begin{align}\label{eqn:apndx-measurement-eqn-2}
  V_{ab}(f) &= \int \mathcal{W}^\textrm{I}_a(\hat{\mathbf{s}},f)\mathcal{W}^{\textrm{I}\star}_b(\hat{\mathbf{s}},f)\,\mathcal{I}(\hat{\mathbf{s}},f)\,e^{-2\pi i f\mathbf{r}_{ab}\!\cdot\,\hat{\mathbf{s}}/c}\,\dif\Omega,
\end{align}
where, superscript `$\textrm{I}$' in the $\mathcal{W}$ term indicates the weighting is inherently introduced by the instrument during the measurement process. The beam-weighted dirty image is:
\begin{align}
  \mathcal{I}_\textrm{D}(\hat{\mathbf{s}},f) &= \frac{1}{\Nant(\Nant-1)}\,\sum_{ab}\, V_{ab}(f)\,e^{2\pi i f\mathbf{r}_{ab}\!\cdot\,\hat{\mathbf{s}}/c},
\end{align}
where, $N_A$ is the number of antennas in the array and we have assumed equal weighting for each antenna pair.

When the array is homogeneous, $\mathcal{W}^\textrm{I} \equiv \mathcal{W}^\textrm{I}_a \equiv \mathcal{W}^\textrm{I}_b$, and the above equation reduces to:
\begin{align}\label{eqn:apndx-dirty-image-homogeneous}
  \mathcal{I}_\textrm{D}(\hat{\mathbf{s}},f) &= |\mathcal{W}^\textrm{I}(\hat{\mathbf{s}},f)|^2\,\mathcal{I}_\textrm{D}^\textrm{iso}(\hat{\mathbf{s}},f),
\end{align}
where, $\mathcal{I}_\textrm{D}^\textrm{iso}(\hat{\mathbf{s}},f)$ is the dirty image with no beam weighting (isotropic, uniform weighting of the sky) determined only by the array layout and $|\mathcal{W}^\textrm{I}|^2$ is the directional power pattern of the antenna pair familiar in standard interferometry.

In order to understand an image from a heterogeneous array, we start by looking at the contribution to the image from the visibility, $V_{ab}$, from each antenna pair. While imaging, EPIC introduces a weighting to each antenna during gridding as given by equations~\ref{eqn:dirty-image-FX} and \ref{eqn:e-field-conv}, and equivalently, the weighted visibility from an antenna pair $ab$ projected on the grid is:
\begin{align}
  V^\prime_{ab}(\mathbf{r},f) &= B^{\textrm{G}\star}_{ab}(\mathbf{r}-\mathbf{r}_{ab},f)\,V_{ab},
\end{align}
where, $B^\textrm{G}_{ab}(\mathbf{r}-\mathbf{r}_{ab},f)$ is obtained by the spatial cross-correlation of weighting kernels associated with the individual antennas using the definition:
\begin{align}
  B^\textrm{G}_{ab}(\mathbf{r}-\mathbf{r}_{ab},f) &= \int \left[ W^{\textrm{G}\star}_b(\mathbf{r}^\prime+\mathbf{r}-\mathbf{r}_b,f)\right. \nonumber\\
  &\qquad\qquad\qquad \left.\times\, W^\textrm{G}_a(\mathbf{r}^\prime-\mathbf{r}_a,f)\right]\,\dif^2\mathbf{r}^\prime.
\end{align}
The superscript `$\textrm{G}$' denotes the weighting introduced in analysis during the gridding process. Optimal imaging requires $W^\textrm{G}_a = W^\textrm{I}_a$ \citep{mor09,mor11}. However, we keep the two superscripts separately to describe output images under the wrong assumption that the array is homogeneous when instead it is actually heterogeneous.

The sky response of the weighted visibility is:
\begin{align}
  \mathcal{I}^\prime_{ab}(\hat{\mathbf{s}},f) &= \int V^\prime_{ab}(\mathbf{r},f)\,e^{2\pi i f\mathbf{r}\cdot\,\hat{\mathbf{s}}/c}\,\dif^2\mathbf{r} \nonumber\\
                                              &= \int B^{\textrm{G}\star}_{ab}(\mathbf{r}-\mathbf{r}_{ab},f)\,V_{ab}\,e^{2\pi i f\mathbf{r}\cdot\,\hat{\mathbf{s}}/c}\,\dif^2\mathbf{r} \nonumber\\
                                              &= V_{ab}\,\int \left[\int W^\textrm{G}_b(\mathbf{r}^\prime+\mathbf{r}-\mathbf{r}_a,f)\right. \nonumber\\
                                              &\qquad\qquad\,\,\, \left.\times\, W^{\textrm{G}\star}_a(\mathbf{r}^\prime-\mathbf{r}_a,f)\,\dif^2\mathbf{r}^\prime\right]\,e^{2\pi i f\mathbf{r}\cdot\,\hat{\mathbf{s}}/c}\,\dif^2\mathbf{r} \nonumber\\
                                              % &\qquad\qquad \times e^{2\pi i f\mathbf{r}\cdot\,\hat{\mathbf{s}}/c}\,\dif^2\mathbf{r} \nonumber\\
                                              &= \mathcal{W}^{\textrm{G}\star}_a(\hat{\mathbf{s}},f)\mathcal{W}^\textrm{G}_b(\hat{\mathbf{s}},f)\,V_{ab}\,e^{2\pi i f\mathbf{r}_{ab}\!\cdot\,\hat{\mathbf{s}}/c},
\end{align}
which denotes the fringe from the weighted visibility. The image output of EPIC is equivalent to averaging these weighted fringes from all antenna pairs:
\begin{align}\label{eqn:apndx-wt-dirty-image-EPIC}
  \mathcal{I}^\prime(\hat{\mathbf{s}},f) &= \frac{1}{\Nant(\Nant-1)}\,\sum_{\substack{ab\\a\ne b}}\,\mathcal{I}^\prime_{ab}(\hat{\mathbf{s}},f) \nonumber\\
                                         &= \frac{1}{\Nant(\Nant-1)}\,\sum_{\substack{ab\\a\ne b}}\,\left[\mathcal{W}^{\textrm{G}\star}_a(\hat{\mathbf{s}},f)\mathcal{W}^\textrm{G}_b(\hat{\mathbf{s}},f)\right. \nonumber\\
  &\qquad\qquad\qquad\qquad\qquad\qquad \times \left. V_{ab}\,e^{2\pi i f\mathbf{r}_{ab}\!\cdot\,\hat{\mathbf{s}}/c}\right].
\end{align}

If all antennas are identically weighted ($W^\textrm{G} \equiv W^\textrm{G}_a \equiv W^\textrm{G}_b$), using equation~\ref{eqn:apndx-dirty-image-homogeneous}, equation~\ref{eqn:apndx-wt-dirty-image-EPIC} reduces to: 
\begin{align}\label{eqn:apndx-wt-dirty-image-homogeneous}
  \mathcal{I}^\prime(\hat{\mathbf{s}},f) &= \frac{\left|\mathcal{W}^\textrm{G}(\hat{\mathbf{s}},f)\right|^2}{\Nant(\Nant-1)}\,\sum_{\substack{ab\\a\ne b}}\,V_{ab}\,e^{2\pi i f\mathbf{r}_{ab}\!\cdot\,\hat{\mathbf{s}}/c} \nonumber\\
  &= \left|\mathcal{W}^\textrm{G}(\hat{\mathbf{s}},f)\right|^2\,\mathcal{I}_\textrm{D}(\hat{\mathbf{s}},f) \nonumber\\
  &= \left|\mathcal{W}^\textrm{G}(\hat{\mathbf{s}},f)\right|^2\,\left|\mathcal{W}^\textrm{I}(\hat{\mathbf{s}},f)\right|^2\,\mathcal{I}_\textrm{D}^\textrm{iso}(\hat{\mathbf{s}},f).
\end{align}
Thus, the dirty image, which is already attenuated by the instrumental power pattern, gets further attenuated by the power pattern introduced in the gridding step in EPIC imager. This is consistent with \citet{mor09}.

We note that all quantities in the celestial plane (calligraphic fonts) are functions of position, $\hat{\mathbf{s}}$, and frequency, $f$. For convenience, we drop writing this dependence explicitly hereafter.

For a heterogeneous array, it is necessary to keep track of antennas and antenna pairs of different types which will determine the final weighting in the synthesized image. We consider a heterogeneous array consisting of a total of $\Nant$ antennas under $N_\textrm{T}$ different types. And under each antenna type $i$, there are $n_i$ antennas such that $\sum_i n_i = \Nant$. The total number of unique non-zero spacing antenna pairs is $\Nant(\Nant-1)/2$ and there are potentially up to $N_\textrm{T}(N_\textrm{T}+1)/2$ unique antenna pair types obtained by pairwise combination of the $N_\textrm{T}$ antenna types. The total number of antenna pairs, 
\begin{align}\label{eqn:apndx-total-baselines}
  \Nant(\Nant-1) &= \sum_{ij} n_{ij}, \\
  \textrm{where}, \quad   n_{ij} &= \begin{cases}
    n_i\,n_j, & i\ne j \\
    n_i(n_i-1), & i=j, 
  \end{cases}
\end{align}
is obtained by counting antenna pairs, $n_{ij}$, in different antenna pair types, $ij$.

Since equation~\ref{eqn:apndx-wt-dirty-image-EPIC} is obtained by averaging weighted fringes over all antenna pairs, it can be expressed as: 
\begin{align}\label{eqn:apndx-wt-dirty-image-EPIC-decomp}
  \mathcal{I}^\prime &= \frac{1}{\Nant(\Nant-1)}\,\sum_{ij}\sum_{\substack{ab\\a\ne b\\a\in i, b\in j}}\,\mathcal{I}^\prime_{ab}
\end{align}
Here, equation~\ref{eqn:wt-dirty-image-EPIC} has been re-written as the sum of fringes over all antenna pairs in an antenna pair type (inner sum) and subsequently summed over all antenna pair types (outer sum). The effective attenuation of the dirty image due to instrumental and gridding weights is:
\begin{align}\label{eqn:apndx-effective-weighting}
  \mathcal{W}_\textrm{eff} &= \mathcal{I}^\prime\, / \,\mathcal{I}_\textrm{D}^\textrm{iso}.
\end{align}

We can simplify further if we assume that the spatial distribution of antenna pairs in each antenna pair type are similar and thus result in a similar dirty image, $\mathcal{I}_\textrm{D}^\textrm{iso}$. This usually holds when $\Nant$ is large and antennas under each antenna type are chosen randomly such that differences in the point spread functions are insignificant. Under this assumption, equation~\ref{eqn:apndx-wt-dirty-image-EPIC-decomp} can be written as:
\begin{align}\label{eqn:apndx-t-dirty-image-EPIC-decomp-approx1}
  \mathcal{I}^\prime &\approx \frac{\mathcal{I}_\textrm{D}^\textrm{iso}}{\Nant(\Nant-1)}\,\sum_{ij}\sum_{\substack{ab\\a\ne b\\a\in i, b\in j}} \mathcal{W}^{\textrm{G}\star}_a\,\mathcal{W}^\textrm{G}_b\,\mathcal{W}^\textrm{I}_a\,\mathcal{W}^{\textrm{I}\star}_b,
\end{align}
where, $\mathcal{I}_\textrm{D}^\textrm{iso}$ is the same for all terms and has been pulled outside the summations. For all antennas indexed by $a$ that belong to a particular type $i$, we replace $\mathcal{W}_a$ with $\mathcal{W}_{(i)}$ {\bf and is applicable to those arising} from both instrumental (superscript `$I$') and gridding (superscript `$G$') origins. Using arguments similar to those in equation~\ref{eqn:apndx-total-baselines}, equation~\ref{eqn:apndx-wt-dirty-image-EPIC-decomp-approx1} can be further simplified to:
\begin{align}\label{eqn:apndx-wt-dirty-image-EPIC-decomp-approx2}
  \mathcal{I}^\prime &\approx \frac{\mathcal{I}_\textrm{D}^\textrm{iso}}{\Nant(\Nant-1)}\,\sum_{ij}n_{ij}\,\mathcal{W}^{\textrm{G}\star}_{(i)}\,\mathcal{W}^\textrm{G}_{(j)}\,\mathcal{W}^\textrm{I}_{(i)}\,\mathcal{W}^{\textrm{I}\star}_{(j)}.
\end{align}

This means $\mathcal{W}_\textrm{eff}$ can be expressed as fourth power combinations of antenna voltage patterns each weighted by the number of antenna pairs in those antenna pair types:
\begin{align}
  \mathcal{W}_\textrm{eff}^\textrm{opt} &\approx \frac{1}{\Nant(\Nant-1)}\,\sum_{ij}n_{ij}\,\mathcal{W}^{\textrm{G}\star}_{(i)}\,\mathcal{W}^\textrm{G}_{(j)}\,\mathcal{W}^\textrm{I}_{(i)}\,\mathcal{W}^{\textrm{I}\star}_{(j)}.
\end{align}
In an optimally weighted image, $\mathcal{W}^\textrm{G}_{(i)} \equiv \mathcal{W}^\textrm{I}_{(i)}$. Then, equation~\ref{eqn:apndx-effective-weighting} reduces to:
\begin{align}\label{eqn:apndx-effective-weighting-optimal}
  \mathcal{W}_\textrm{eff}^\textrm{opt} &\approx \frac{1}{\Nant(\Nant-1)}\,\sum_{ij} n_{ij}\,\left|\mathcal{W}^\textrm{I}_{(i)}\right|^2\left|\mathcal{W}^\textrm{I}_{(j)}\right|^2,
\end{align}
where the superscript `$\textrm{opt}$' denotes `optimal'.}
